Fast and adaptive cointegration based model for forecasting high frequency financial time series

### 2016-07-06 articles-conferences/iscef2016/src/background.tex
manuscript updates

\def\rot#1{{\color{red}{#1}}}
\def\gruen#1{{\color{green}{#1}}}
\def\blau#1{{\color{blue}{#1}}}

@@

\subsection{Integration and Cointegration}\label{sec:coint}\  
Following Johansen \citep{johansen1995} we shall say that a stochastic process
$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
$\varepsilon_t\sim iid(0,\sigma^2)$ holds.

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is
obtained \citep{banerjee1993}; more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

Vector error correction model (VECM) describe the joint behaviour of a set of
variables and can be derived from the simple Vector Autoregressive model (VAR)
\citep{sims1980} model.  The VAR($p$) model is a framework describing the
behaviour of a set of $l$ endogenous and stationary variables as a linear
combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

@@

However, the VAR model cannot be used with non-stationary variables. VECM
\citep{engle87} is also a linear model and a special form of a VAR model for I(1)
variables that are also cointegrated \citep{banerjee1993}. If cointegration
exists, variable differences are stationary and they introduce an error
correction term which adjusts coefficients to bring the variables back to
equilibrium.

It is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\citep{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\citep{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{B}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\label{eq:Amatrix}
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares (OLS) estimation.

>>>>

@@

\subsection{Integration and Cointegration}\label{sec:coint}\  
Following \cite{johansen1995} we shall say that a stochastic process
$Y_t$ which satisfies $Y_t-E(Y_t) = \sum_{i=0}^\infty C_i\,\varepsilon_{t-i}$ is
called $I(0)$, and then we shall write $Y_t\sim I(0)$, whenever
$\sum_{i=0}^\infty C_i \neq 0$ and $\sum_{i=0}^\infty C_i\,z^i$ converges for
$z\in\mathbb{C}$ with $|z|<1$.  It is understood that the condition
$\varepsilon_t\sim iid(0,\sigma^2)$ holds.

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation an stationary process is
obtained, see \cite{banerjee1993}; more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

Vector error correction model (VECM) describe the joint behaviour of a set of
variables and can be derived from the simple Vector Autoregressive model (VAR)
presented by \cite{sims1980}.  The VAR($p$) model is a framework describing the
behaviour of a set of $l$ endogenous and stationary variables as a linear
combination of their last $p$ values, where $l,p\in\mathbb{N}$.  In our case,
each one of these $l$ variables is a scalar time series $y_{\lambda,t}$,
$\lambda=1,\dots,l$, and we represent them all together at time $t$ by the
vector time series:
\begin{equation}
\label{eq:variables}
\mathbf{y}_t = 
\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t} \end{bmatrix}^\top.
\end{equation}
\noindent
Notice that the vectors $\mathbf{y}_t$ are assumed to be $l$-dimensional.

@@

However, the VAR model cannot be used with non-stationary variables. VECM ,
developed by \cite{engle87}, is also a linear model for I(1) variables that are
also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium.

It is obtained re-writing equation (\ref{eq:var}) in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of (\ref{eq:var}), are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$, see
\cite{johansen1995}, will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure,see \cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
non-singular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

Transposing each equation of the system (\ref{eq:vecfull}) we can write
the VECM($p$) model in block-matrix form as:
\begin{equation}\label{eq:vareq}
\mathbf{B} = 
\mathbf{A} \mathbf{X} + 
\mathbf{E} \, , 
\end{equation}
%
\noindent where $\mathbf{B}$ dimension is $((N-p)\times l)$, $\mathbf{A}$
dimension is $((N-p)\times(r+(p-1)l +1))$, $\mathbf{X}$ dimension is $((r+(p-1)l
+1)\times l)$ and $\mathbf{E}$ dimension is $((N-p)\times l)$:
%
\begin{alignat}{3}
\mathbf{B}
&= \begin{bmatrix}
   \Delta\mathbf{y}_{p+1}^\top \\
   \Delta\mathbf{y}_{p+2}^\top \\
   \vdots \\
   \Delta\mathbf{y}_N^\top
   \end{bmatrix}
&\quad
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\alpha}^\top \\
   \boldsymbol{\Phi}_1^{*\top} \\
   \boldsymbol{\Phi}_2^{*\top} \\
   \vdots \\
   \boldsymbol{\Phi}_{p-1}^{*\top} \\
   \mathbf{c}^\top
   \end{bmatrix}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}
\end{alignat}
\noindent and 
\begin{align}
\mathbf{A} 
&= \begin{pmat}[{....|}]
   \mathbf{y}_p^\top \boldsymbol{\beta} & \Delta \mathbf{y}_p^\top & \Delta\mathbf{y}_{p-1}^\top & \dots 
                    & \Delta\mathbf{y}_2^\top & 1 \cr
   \mathbf{y}_{p+1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{p+1}^\top & \Delta\mathbf{y}_p^\top & \dots
                       & \Delta\mathbf{y}_3^\top & 1 \cr
   \vdots & \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top  \boldsymbol{\beta} &\Delta\mathbf{y}_{N-1}^\top & \Delta\mathbf{y}_{N-2}^\top & \dots 
                       & \Delta\mathbf{y}_{N-p-1}^\top & 1 \cr
   \end{pmat}\, .
\label{eq:Amatrix}
\end{align}
Taking into account the error term $\mathbf{E}$, equation~(\ref{eq:vareq}) 
can be solved with respect to $\mathbf{X}$ using the OLS estimation.

### 2016-07-06 articles-conferences/iscef2016/src/conclusions.tex
Final manuscript
manuscript updates

\section{\uppercase{Conclusions}}
\label{sec:conclusions}
Cointegration in financial time series has been largely studied and the 
Johansen method is commonly used to obtain cointegration relationships. 
In practice, it has been found that cointegration relations change with time. 
However, model-based cointegration such as VECM assumes that cointegration
remains unchanged in time. 
We empirically showed that the Johansen method is sensitive to the number
of lags but also to the amount of data considered.

@@

Determining VECM parameters was the most expensive routine and it was run
using parallel processes using MPI which allowed a grid search within a range of
values for $L$ and $p$ to be made.
Tests were done using real currency rates data. 
Our proposal was tested in several scenarios using different times of
the day to show its behaviour related to the opening times of
financial markets.

Results showed that our proposed AVECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
These facts were consistent for different times of the day. 
We found that increasing the number of parameters will always lead to
better performance measures.

The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds. Since we used 10-second frequencies we can say that our proposal is suitable
for use in an online context for real applications because response times
were less than this frequency.

>>>>

\section{Conclusions}
\label{sec:conclusions}
Cointegration in financial time series has been largely studied and the Johansen
method is commonly used to obtain cointegration relationships.  In practice, it
has been found that cointegration relations change with time.  However,
model-based cointegration such as VECM assumes that cointegration remains
unchanged in time.  We empirically showed that the Johansen method is sensitive
to the number of lags but also to the amount of data considered.

@@

Determining VECM parameters was the most expensive routine and it was run using
parallel processes using MPI which allowed a grid search within a range of
values for $L$ and $p$ to be made.  Tests were done using real currency rates
data.

Results showed that our proposed AVECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.

The parallel implementation allowed the execution times to be reduced more than
9 times and therefore a response time was obtain before 10 seconds. Since we
used 10-second frequencys we can say that our proposal is suitable for use in
an online context for real applications because response times were less than
this frequency.

### 2016-07-06 articles-conferences/iscef2016/src/intro.tex
manuscript updates

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that
some sets of variables cannot wander too far from each other. Cointegration
means that one or more linear combinations of these variables are stationary
even though individually they are not. Some models,
such as the Vector Error Correction (VECM) \cite{engle87}, take advantage of
this property and describe the joint behaviour of several cointegrated
variables. VECM introduces this long-run relationship among a set of cointegrated
time series as an error correction term. These time series must be integrated of order 1, denoted I(1), i.e. they become stationary at their first differences. In finance, I(1) time series are very common and 
to introduce cointegration restrictions in models often improves forecasting \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications \cite{mukherjee1995,seong2013,maysami2000,arestis2001} such as pair trading \cite{herlemont2003} or models with more than two variables \cite{mukherjee1995,engle2004}.

VECM parameters are obtained using the ordinary least squares
(OLS) method \cite{golub1980}. Since OLS involves many calculations, the parameter estimation is computationally expensive when the number of lagged values and
data increases. Moreover, obtaining cointegration vectors 
is also an expensive routine. This is a main limitation to use VECM with stream data with high frequency.
Chen and Lung's BVECM \cite{chen2003} addresses the advantage of distributed processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high frequency stream data.
Our approach was to determine adaptively the number of observations and lags of VECM which
maximise cointegration relations in the past. 
 This search is computationally
expensive because the Johansen
method \cite{johansen1995} is required to find cointegration vectors, which is of order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error Correction (AVECM), 
is to parallelise this search in order to get new parameters before new data arrives.
Model effectiveness is focused on out-of-sample forecast rather than in-sample
fitting. This criterion allows AVECM prediction capability to be
expressed rather than just explaining data history.
The forecast capability of our method was measured using MSE and the Theil's $U$-statistic \cite{theil1966} widely used in economic forecast. Tests were run using four currency rates: Euro (EUR) to
United States Dollar (USD) (EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss
Franc (CHF) (USDCHF) and USD to Japanese Yen (JPY) (USDJPY) with a 10-seconds frequency.

>>>>

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM), see \cite{engle87}, take advantage of this property and describe the
joint behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}).

VECM parameters are obtained using the ordinary least squares (OLS) method,
developed by \cite{golub1980}. Since OLS involves many calculations, the
parameter estimation is computationally expensive when the number of lagged
values and data increases. Moreover, obtaining cointegration vectors is also an
expensive routine. This is a main limitation to use VECM with stream data with
high frequency. \cite{chen2003} addresses the advantage of distributed
processing over conventional rolling window processing.

Our aims were to study if a parallel version of VECM can be used with high
frequency stream data.  Our approach was to determine adaptively the number of
observations and lags of VECM which maximise cointegration relations in the
past.  This search is computationally expensive because the Johansen method, see
\cite{johansen1995}, is required to find cointegration vectors, which is of
order $O(n^3)$. Therefore, our proposal, called Adaptively Vector Error
Correction (AVECM), is to parallelise this search in order to get new parameters
before new data arrives.  Model effectiveness is focused on out-of-sample
forecast rather than in-sample fitting. This criterion allows AVECM prediction
capability to be expressed rather than just explaining data history.  The
forecast capability of our method was measured using MSE and the Theil's
$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
were run using four currency rates: Euro (EUR) to United States Dollar (USD)
(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

### 2016-07-06 articles-conferences/iscef2016/src/methodology.tex
manuscript updates

\section{\uppercase{Methodology}} \label{sec:methodology}
\subsection{Motivation} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, the long-run relationships between the time
series might change due to several economic factors that can lead to structural
breaks in the cointegration relationship \citep{gregoryETal1996}.  In order to
show the number of cointegration vectors which depends on the amount $L$ of
historical data and the number of lags $p$ in the VECM, we used a grid search.
We defined a grid of possible values for $L$ and $p$. $L$ goes throughout
$[2,14]$ hours with a step size of 4 hours and $p$  throughout $[1,5]$ with step
size of 1. The first experiment consisted in determining the number of
cointegration vectors for all combinations of $L$ and $p$.  We used four forex
rates: EURUSD, GBPUSD, USDCHF and USDJPY with 10-second frequency. Data started
at 13:00 GMT of the 13th of August 2014, when the New York and London financial
markets opened.

Figure~\ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$ and
$p=1$ . This procedure was carried out by a sliding window of historical data
moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{img/Fig1}
  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
  possible values for $L$ are shown (2, 6, 10 and 14) hours (1 hour = 360 data
  points).}
  \label{fig:hists}
\end{figure}

@@

The goal of our next experiment was to find a relation between this ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[700,1500]$ data points with step size 100 and $p$ takes values between
$[1,5]$ with step size 1.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.8\textwidth]{img/Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best $L$ found was 1060. Below MSE versus $L$ shows a rapidly
  decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

@@

Our proposal was then a modified version of VECM, with parameter $p$ and the
amount of historical data used $L$ obtained at every step. Only the search of
$L$ and $p$ was done in a distributed environment, since this is the most
expensive routine. Our proposal is called Adaptive Vector Error Correction model
(AVECM). AVECM is detailed in the algorithm \ref{alg:AVECM} which summarises
our proposal.

@@

\subsection{Model comparison} \label{sec:random}
We compared our proposal, in terms of performance, with the naive forecast of
the random walk model and ARIMA. It is still difficult to outperform the random
walk model for standard econometric forecasting models \citep{lo2011} despite its
simplicity. The random walk model is defined as:

@@

On the other hand, ARIMA is widely used to forecast returns in finance
\citep{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns,
we used $d=1$.

@@

Forecast performance was evaluated using two different methods which are
frequently used in finance:
\begin{description}
\item
{\bf MSE},  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item {\bf $U$-statistic}, the Theil's $U$-statistic \citep{theil1966} is a unit
free measure obtained as the ratio between the root MSE (RMSE) of a model and
the RMSE of the naive random walk model. A $U$-statistic less than 1 implies the
performance is better than the naive model.
\end{description}

>>>>

\section{Methodology} \label{sec:methodology}
\subsection{Motivation} \label{sec:proposal}

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
long-run relationships between the time series  might change due to several
economic factors that can lead to structural breaks in the cointegration
relationship.  In order to show that the number of cointegration vectors depends
on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
consisted in determining the number of cointegration vectors for all
combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
August 2014, when the New York and London financial markets opened.

Figure \ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$
hours  and $p=1$ . This procedure was carried out by a sliding window of
historical data moving 1000 times.

\begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
  \includegraphics[width=0.9\textwidth]{Fig1}
  \caption{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}

@@

The goal of our next experiment was to find a relation between this ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[700,1500]$ data points with step size 100 and $p$ took values between
$[1,5]$ with step size 1.

\begin{figure}[ht!]
  \centering
  \includegraphics[width=0.9\textwidth]{Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

@@

Our proposal is then a modified version of VECM, with parameter $p$ and the
amount of historical data used $L$ obtained at every step. Only the search of
$L$ and $p$ was done in a distributed environment, since this is the most
expensive routine. Our proposal is called Adaptive Vector Error Correction model
(AVECM). AVECM is detailed in the algorithm \ref{alg:AVECM} which summarises
our proposal.

@@

\subsection{Model comparison} \label{sec:random}
We compared our proposal, in terms of performance, with the naive forecast of
the random walk model and ARIMA. It is still difficult to outperform the random
walk model for standard econometric forecasting models  despite its
simplicity, see \cite{lo2011}. The random walk model is defined as:

@@

On the other hand, ARIMA is widely used to forecast returns in finance, see
\cite{tsay2005}. A process can be modelled as an ARIMA$(p,d,q)$ model if
$\mathbf{x}_t=\Delta^d \mathbf{y}_t $, i.e after differencing $d$ times the time
series $\mathbf{y}_t$,  we get an ARMA$(p,q)$. Since we are modelling returns,
we used $d=1$.

@@

Forecast performance was evaluated using two different methods which are
frequently used in finance:
\begin{description}
\item
{\bf MSE},  Mean Square Error measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}
\item {\bf $U$-statistic}, the Theil's $U$-statistic, presented by
\cite{theil1966}, is a unit free measure obtained as the ratio between the root
MSE (RMSE) of a model and the RMSE of the naive random walk model. A
$U$-statistic less than 1 implies the performance is better than the naive
model.
\end{description}
